# Counterfactual Model

```{r, echo = FALSE}
library(dplyr)
set.seed(777)
```

## Conterfactual Model as a Missing Data Problem

The counterfactual model assumess that there is an unseen (missing) outcome that our experimental unit would have displayed had it received a different treatment than it did. The following data example will be useful to understand this point. We have $n =6$ units that randomly get assigned a treatment $A$ like so:

```{r}
n_obs <- 6
obs_id <- 1:n_obs
A <- rbinom(n_obs, 1, 0.5)
```

We denote by $Y$ their individual responses to the treatment. In this example, receiving the outcome increases on average its baseline of 100 by 10 units. In statistical language we can write this as

\begin{equation}
Y=100+10A+\varepsilon   (\#eq:assignment)
\end{equation}

where $\varepsilon$ is normally distributed error. This can be simulated as:

```{r}
epsilon <- rnorm(n_obs)
Y <- 100 + 10*A + epsilon
```

This sort of data would be represented like this in a typical data set:

```{r}
data.frame(obs_id, A, Y)
```

However, this data generating mechanism in \@ref(eq:assignment) can be thought in a different way as well.

\begin{eqnarray}
Y(0) = & 100 +\varepsilon, \\
Y(1) = & 110 + \varepsilon. (\#eq:conterfactual)
\end{eqnarray}

The two equations in \@ref(eq:conterfactual) say that each experimental unit would express one of the two outcomes 

\begin{equation}
(Y(1), Y(0)) (\#eq:potentialoutcomes)
\end{equation}

depending on the treatment $A$ assigned. Only one of the two potential outcomes in \@ref(eq:potentialoutcomes) would be observed.

In a world where we could know what all the conterfactual outcomes were we would collect data as follow:

```{r}
Y_0 <- 100 + epsilon
Y_1 <- 100 + 10*A + epsilon
```

where we can record both $Y(0)$ and $Y(1)$ at the same time on each experimental unit. Then our data would look like

```{r}
data.frame(obs_id, A, Y_0, Y_1)
```


but we don't live in a world with parallel universes so in reality we would only have this

```{r}
data.frame(obs_id, A, Y_0, Y_1) %>%
   mutate(
      Y_0 = ifelse(A == 0, Y_0, NA),
      Y_1 = ifelse(A == 1, Y_1, NA)
   )
```

and by a minor data wrangling we get back to our original data:

```{r}
data.frame(obs_id, A, Y_0, Y_1) %>%
   mutate(
      Y_0 = ifelse(A == 0, Y_0, NA),
      Y_1 = ifelse(A == 1, Y_1, NA),
      Y   = ifelse(A == 1, Y_1, Y_0)
   ) %>% 
  select(obs_id, A, Y)
```

## Chapter Summary

